Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-18 paper.tex
Pass through recombination parents section

\subsection{Divergence between recombinant parents}
\label{sec:parent_divergence}
% - Divergence between recombinant parents
%     We study the recombination events in the Long ARG in terms. A recombinant
% will not always have two parents, for reasons X, Y and Z. As a first
% approximation, we can assume that each breakpoint corresponds to a
% recombination event (although there are some exceptions). Here we have to get
% into all the hairy details about which sets of breakpoints we filter and don't
% filter.

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas}
Date of common ancestry between the parents on either side of recombination
breakpoints, as a function of parental divergence. MRCAs of parents
inferred to have caused a Pango designated recombinant (XA, XB, etc.)
identified in orange (A) 851 parent pairs (singleton recombinants removed).
Parental strains often descend from the same MRCA; each horizontal dotted line
indicates a frequently found MRCA: these tend to be associated with major waves
and often have many immediate children (i.e. are large
polytomies, likely to indicate undersampled outbreaks) (B) Further filtered
to 358 parent pairs by removing HMM inconsistent breakpoints. For equivalent
plots broken down by parental VoC classification, see
Figure~\ref{fig:recomb_mrcas_voc_breakdown}
}
\end{figure}

Although recombination nodes are relatively infrequent, there are still many
present in the wide and long ARGs (4123 and 2078 respectively; Supplementary
Tables 1 and 2). Each recombination node was inferred as a
result of matching a specific (``causal'') recombinant sample into the ARG.
However, ~63\% of them are ancestral to only a single
recombinant sample; we exclude these "singleton recombinants" from subsequent
analyses as they may have arisen from sequencing of mixed samples (e.g., from
co-infection or contamination), and in any case, never result in onward
transmission.  This leaves 1522 and 763 recombination nodes in the wide and
long ARGs, respectively. In their analysis using RIPPLES,
\cite{Turakhia2022-it} detected 589 non-singleton recombinants.

We focus on the long ARG because compared to the wide ARG, it includes time
periods of increased rates of coinfection and recombination, and contains many
samples previously identified as the product of recombination (i.e. with a
Pango lineage designation starting with an "X"). Not only does the long ARG
capture all but one of the Pango designated recombinant lineages during this
time period (see the section below), but it also includes a large number of
non-Pango-identified recombinants. Many of these involve recombination between
relatively closely related lineages, as measured by the divergence time between
the recombination events and the most recent common ancestor (MRCA) shared by
the left and right parent at each breakpoint. As can be seen from
Figure~\ref{fig:recomb_mrcas}, this divergence time ranges from as little as
five or six days, to as much as several years in the case of recent
recombinants. This demonstrates that sc2ts can identify viral recombination
between very closely related lineages, providing a degree of resolution not
available with other recombination detection methods.

The histograms in Figure~\ref{fig:recomb_mrcas} show that
majority of newly identified recombinants arise from parents that diverged 10 to 30 weeks previously.
However, there is also a peak of recombinants that diverged ${\sim}95$ weeks prior to the event:
these correspond to recombinants whose parental common ancestor traces to early 2020,
including both Delta-Delta and Delta-Omicron recombinants
(Supplementary Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
Note that each plotted point represents a breakpoint, so that a recombination
event with more than one breakpoint (e.g. with 3 or more parents, comprising
${\sim}10 \%$ of the recombination events in Figure ~\ref{fig:recomb_mrcas}A and
${\sim}3 \%$ in Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.

As a check on the reliability of these identified recombinations,
Figure~\ref{fig:recomb_mrcas}B  plots only breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes \&
breakpoints in the long ARG from 763 \& 851 to 347 \& 358
(and in the wide ARG from 1522 \& 1769 to XXX \& YYY , see
Supplementary Tables 1 and 2).

The date of common ancestry for the two parents of recombinant lineages is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a small handful of MRCAs which are shared between a large number
of recombination events; the top four are indicated by dotted lines in the
figure. These shared MRCAs lie near the root of large expansions of certain
lineages (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
and BA.2: second Omicron wave), as evidenced by the number of samples with those
Pango designations which trace back to each ancestral node (listed as
percentages on the plot). The majority of these frequently-shared MRCAs
are associated with extensive polytomies in the tree (at the extreme,
the BA.2-linked node has 2,872 immediate children). This is likely to
indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.

Finally, it is clear from Supplementary Figure\ref{fig:recomb_mrcas_voc_breakdown} that of the recombinant
parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
are within-Delta and within-Omicron recombinants, and that Omicron followed by
Delta, are the variants associated with the most recombination. This may
reflect either sampling intensity, the prevalence of cases (which increases the
chance of coinfection and recombination), or possibly heterogeneity in
recombination probabilities among lineages.

@@

The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node is ancestral to more than
127K samples, and is likely to be a spuriously inferred recombination event.
Understanding the origin of this and other similar nodes, is an important
direction for future work (see Discussion). For simplicity, we exclude these
samples from further analyses in this section.

@@

\caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node are shown
corresponding to the maximum inferred breakpoint position from the HMM process.
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

@@

Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type with no other Pango designations among their descendants.
The remaining 10 are of the XAG type with multiple Pango
designated lineages as additional descendants of the originating recombination event. In some cases,
these may be, however, be a result of erroneous Pango designations. Supplementary Table~\ref{tab:pango-recombinants} also shows the official Pango designated parent lineages for
each recombinant lineage. The parent lineages in the Long ARG agree to a great extent with these, although \texttt{sc2ts} provides more a precise
parent designation.

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. 15 Pango X lineages are inferred to have multiple
recombinant origins, such that their samples trace to >1 most recent
recombination node in the ARG. Of these, 11 are cases where none of the
recombinants are nested (i.e. there is no overlap between the descendant samples
of each recombination node). In 9 of these 11, there is a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend (in other two, there are too few Pango X samples to decide on a ``main'' clade:
see Supplementary Table~\ref{tab:pango-recombinants}).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).

A similar pattern is seen for several of the origins inferred for the Pango
XN lineage (originally defined on the basis of sequences from the UK). The
Long ARG infers four independent origins for nonsingleton Nextclade-designated XN
samples (Supplementary Table~\ref{tab:multiple_origins_table}). One of the recombinant origins,
caused in our ARG by an English sample, accounts for 155 out of 164 XE samples
(primarily UK and Irish, but with a few northern European and north American strains).
This certainly represents the XE recombinant lineage as originally defined. The other recombination nodes
only give rise for one, two, or three XE samples. Three of these recombination
nodes are likely to represent Nextclade mislabelling: they either have different
inferred breakpoint positions or (in the case of the earliest recombination node),
descendants that are primarily labelled as other Pango X-lineages. Additional support
for the suggestion that they result from incorrect labelling comes from the fact that
none of their XE-labelled descendant samples are from the UK or Ireland.
It is less clear if the most recent XE cluster is a mislabelling or a case of \texttt{sc2ts}
inferring multiple recombinations when there should be just one. This cluster of 2 samples
(one of which is English) has exactly the same breakpoint estimates as the
main cluster. If the only samples available are rather genetically distant from an originating
recombinant, we might expect sc2ts to infer multiple similar recombination events
(rather than a single event with extra reversions)

@@

As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked with asterisks in Supplementary Table~\ref{tab:pango-single-origin}; the most
extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the Wide ARG (but as discussed previously,
in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
descendants).

@@

\begin{table}
\small
\centering
\begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
\toprule
Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
\midrule
2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
\bottomrule
\end{tabular}
\caption{\label{tab:multiple_origins_table}
Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
(Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
XN samples are grouped into rows based on their most recent ancestral recombination node;
Details are also provided for the ``causal strain'' that led to the recombination
being proposed by \texttt{sc2ts}. Note that a large number of XN samples were previously removed by filtering. When undersampling causes three or more mutations to lie immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple independent origins rather than one.
}
\end{table}

>>>>

\subsection{Divergence between recombinant parents}
\label{sec:parent_divergence}
% Paragraph 1, what does all this mean?

% Purpose of this paragraph - remind the reader that we can only do this
% sort of things because we have an ARG, and we can do  this at a far
% greater level of granularity than existing methods.
[ FIXME this is a rough outline of what an intro para might say]
An important advantage of inferring a joint model of inheritance,
mutation and recombination in an ARG (Section~\ref{sec:args})
is that is allows us to precisely trace the origins of recombinants.
[Katherine - need some input from you here to say this properly?]
Existing methods only provide us with approximate estimates of
the origins of recombination events, and are not able to distinguish
recombinations between closely related lineages.

% Paragraph 2, what do we plot, specifically?
In this section we investigate the patterns of divergence between
the parents of recombination nodes.
We focus on the Long ARG because  it covers time
periods where recombination is known to have had a significant
impact, and includes
[HOW MANY?]
samples designated to [HOW MANY] Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis of Pango recombinant lineages in the Long and Wide ARGs).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e., the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, we exclude recombination nodes
that are ancestral to only one sample from this analysis.
Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all recombinants that
are ancestral to more than one sample, and
Figure~\ref{fig:recomb_mrcas}B plots a subset which pass some further
reliability filters. Specifically, only recombination nodes that
[FILL ME IN].
This reduces the number of recombination nodes
to 347 and breakpoints to 358.
Note that each plotted point
represents a breakpoint, so that a recombination node with more than one
breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
It is important to note here that time estimates in these figures may be unreliable,
since nodes other than samples are dated somewhat crudely in the current version
of \texttt{sc2ts} (Section~\ref{sec:node_dating}).

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas}
Date of common ancestry between the parents on either side of recombination
breakpoints, as a function of the divergence time between the parents.
MRCAs of parents associated with Pango designated recombinants (XA, etc)
are identified in orange. (A) 851 parent pairs (singleton recombinants removed).
(B) Further filtered
to 358 parent pairs by removing HMM inconsistent breakpoints (see text).
Horizontal dotted lines show the four most common MRCA nodes,
which tend to be associated with major outbreaks and with many immediate children.
See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
for equivalent plots broken down by parental VoC classification.
}
\end{figure}

% Paragraph 3, what do we say about the dates/nodes of the MRCAs?
The date of the MRCA of recombinant parents is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a few MRCAs shared by many
recombinants; the top four are indicated by dotted lines in the
figure. These shared MRCAs lie near the root of large expansions
(B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
and BA.2: second Omicron wave), as demonstrated by the number of samples with those
Pango designations which trace back to each ancestral node (listed as
percentages on the plot). The majority of these frequently-shared MRCAs
are associated with large polytomies (at the extreme,
the BA.2-linked node has 2,872 immediate children), likely
indicating a rapid and under-sampled expansion of a clade.

% Paragraph 4, what do we say about the divergence times?
The histograms in Figure~\ref{fig:recomb_mrcas} show that there is a
broad range in divergance times among recombinant parents.
We can also see that the majority of recombinants arise
from parents that diverged 10 to 30 weeks
previously. However, there is also a peak of recombinants that diverged
${\sim}95$ weeks prior to the event: these correspond to recombinants whose
parental common ancestor traces to early 2020, including both Delta-Delta and
Delta-Omicron recombinants
(Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
Figure~\ref{fig:recomb_mrcas_voc_breakdown} also shows that, of the recombinant
parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
are within-Delta and within-Omicron recombinants, and that Omicron and
Delta are the variants associated with the most recombination. This may
reflect either sampling intensity, the prevalence of cases (which increases the
chance of coinfection and recombination), or possible heterogeneity in
recombination probabilities among lineages.

@@

The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain Denmark/DCGC-281594/2021,
is ancestral to more than 127K samples, most of which are not designated as
recombinants by any other method. It is highly likely to be a spuriously inferred recombination event.
Understanding the origin of this and other similar nodes is an important
direction for future work (see Discussion). For simplicity, we exclude these
samples from further analyses in this section.

@@

\caption{\label{fig:pango-simple-origin-graph} Non-nested
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node are shown
corresponding to the maximum inferred breakpoint position from the HMM process.
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

@@

Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
basic (XA) type with no other Pango designations among their descendants.
The remaining ten are of the XAG type with multiple Pango
designated lineages as additional descendants of the originating recombination event.
In some cases,
these may be, however, be a result of erroneous Pango designations.
Supplementary Table~\ref{tab:pango-recombinants} also shows the official
Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
which extensively agree (although \texttt{sc2ts} provides more a precise
parent designation).

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. 15 Pango X lineages are inferred to have multiple
recombinant origins, such that their samples trace to >1 most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e. there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend (but in XJ and XU there are too few Pango X samples to decide on a ``main'' clade:
see Supplementary Table~\ref{tab:pango-recombinants}).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).

A similar pattern is seen for several of the origins inferred for the Pango
XN lineage (originally defined on the basis of sequences from the UK). After removing
singleton recombinants and the descendants of the erroneously recombinant
Denmark/DCGC-281594/2021 strain (which, however, includes 53 XN-designated samples)
the Long ARG contains four independent origins for Nextclade-designated XN
samples (Supplementary Table~\ref{tab:multiple_origins_table}). The oldest
of these four origins has 3 mutations above the sample that caused the
recombination to be inferred. It therefore seems likely that we have not managed
to sample an XN strain that is very close to the true recombinant. This may
lead to `texttt{sc2ts} mistakenly inferring multiple recombination events when there

If the only samples available are rather genetically distant from an originating
recombinant, we might expect sc2ts to infer multiple similar recombination events
(rather than a single event with extra reversions)

@@

As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the Wide ARG (but as discussed previously,
in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
descendants).

@@

\begin{table}
\small
\centering
\begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
\toprule
Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
\midrule
2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
\bottomrule
\end{tabular}
\caption{\label{tab:multiple_origins_table}
Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
(Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
The 13 XN samples are grouped into rows based on their most recent ancestral recombination node;
all the descendant samples of the four recombination nodes are designated XN by Nextclade.
Details are also provided for the ``causal strain'' that led to the recombination
being proposed by \texttt{sc2ts}. When undersampling causes three or more mutations to lie
immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple
independent origins rather than one. Note that a large number of XN samples were previously
removed by filtering. 
}
\end{table}
